---
title: "Analysing Mathematical Sciences REF results"
author: "David Selby"
date: "`r format(Sys.Date(), '%e %B %Y')`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Vignette Title}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  message = FALSE
)

```

```{r}
library(ref2014)
```

There were 36 Units of Assessment (UoAs) in the REF.
The names of each of these subject areas are stored in the vector `UoA`.

```{r, results = 'asis'}
data(UoA)
cat(paste('1.', UoA), sep = '\n')
```

This makes it easy to look up the ID of a particular unit of assessment.

```{r}
which(UoA == 'Mathematical Sciences')
grep('Math', UoA)
```

Or to select several related subject areas.

```{r}
grep('Engineering', UoA, value = TRUE)
```

```{r}
library(dplyr)
outputs <- tidy_outputs()
maths_results <- subset(results, `Unit of assessment number` == 10)
maths_outputs <- subset(outputs, uoa_number == 10) %>%
  mutate(row_id = row_number())
```

Key features in the submission data include the titles and DOIs of the research outputs.
In the mathematical sciences data, there `r round(100 * mean(!is.na(maths_outputs$doi)), 2)`% of entries contain DOIs and `r round(100 * mean(!is.na(maths_outputs$issn)), 2)` have ISSNs.

```{r}
## If any two entries share an ISSN, DOI or (standardised) volume title,
## then they MUST be collected into the same journal. 

# maths_long <- maths_outputs %>%
#   select(row_id, doi, volume_std, issn, isbn) %>%
#   tidyr::gather('identifier', 'value', -row_id) %>%
#   filter(!is.na(value)) # Important to avoid grouping by missingness

# library(igraph)
# connected_components <- maths_long %>%
#   select(-identifier) %>%
#   graph_from_data_frame %>%
#   clusters %>% membership %>% stack %>%
#   rename(journal_id = values, row_id = ind)

# maths_outputs2 <- maths_outputs %>%
#   merge(connected_components, by = 'row_id', all.x = TRUE)

maths_outputs2 <- cluster_outputs_by_journals(maths_outputs)

warning('Current implementation is fragile wrt column names')
```

After grouping submissions into journals, we have reduced the dimensionality of the data from `r nrow(maths_outputs)` papers to `r n_distinct(maths_outputs2$journal_id)` journals.
There are `r sum(is.na(maths_outputs2$journal_id))` submissions (`r round(100 * mean(is.na(maths_outputs2$journal_id)), 2)`%) that could not be assigned to a journal as they were missing volume title, DOI, ISSN and ISBN.
Of these unassigned submissions, `r with(maths_outputs2, sum(is.na(journal_id) & output_type == 'D'))` were classified as journal articles.
